Cultural evolutionary dynamics with prestige bias

### 2023-01-04 ms/ms.tex
revise introduction for better definitions of succes prestige influence and motivation for these definitions

Cultural transmission may combine vertical transmission, in which parents transmit to their offspring; oblique transmission, in which adults (teachers, leaders, strangers) transmit traits to unrelated offspring; and horizontal transmission, in which peers from the same age cohort transmit to one another \citep{transmissionVectorsBook,transmissionVectors}.
It has been demonstrated that non-vertical cultural transmission can maintain maladaptive traits, which can be beneficial in changing environments \citep{changeEnv}.

Transmission biases may cause a cultural trait to have a disproportionate probability to be transmitted compared to its frequency in the population.
Importantly, even when a trait is disfavored by natural selection, it may still spread in a population due to transmission biases that are strong enough to overcome selection \citep[Ch.~8, pg.~279]{evolutionBook}. 
For example, \citet{cooperation} have shown that helping behaviors can evolve due to horizontal transmission bias even without any benefit to the recipient, or when the benefit is much larger than the cost.

A common bias in the literature is \textit{success bias}, in which individuals prefer to copy from role models that demonstrate success in some activity, such as fishing, growing yams, and using medicinal plants \citep{fijian_social_bias}, and hunting \citep{Mesoudi2008}, and it can increase the probability of learning a successful trait or strategy \citep{Borofsky2022}. 
Indeed, in a tournament between learning strategies conducted by \citet{strategiesPaper}, most winner strategies included a mixture of success-biased social learning and individual learning, implying that success-biased learning is a good strategy, but that by itself it is not enough to best other strategies, even when success is measured accurately.

\citet[Ch.~5]{evolutionBook} have suggested that the evaluation of success can be divided into three groups: \emph{direct bias}, \emph{indirect bias} and \emph{frequency-dependent bias}.
A direct bias occurs when one phenotype is more attractive than other phenotypes, and is evaluated by \emph{directly} testing the trait.
For example, an individual observing a ping-pong match can attempt the observed paddle grips to determine which grip is better.
An indirect bias occurs when an individual uses one phenotype to determine the attractiveness of another phenotype.
For example, an observer may copy the paddle grip of the ping-pong player who scored more points in the match, thus indirectly evaluating the grip by the points scored.
This, however, may cause mismatches between the copied trait and the rest of the cultural or genetic repertoire of the individual \citep{Kolodny2022}.
A frequency-dependent bias occurs when the probability to copy a phenotype is higher or lower than the frequency of the phenotype among demonstrators. 
For example, suppose the common paddle grip is used by 60\% of the demonstrators; if the this grip is adopted by 80\% of copiers, then transmission is under positive frequency bias, also called \emph{conformity}; if it is adopted by 40\% of copiers, then transmission is under negative frequency bias, or \emph{non-conformity} \citep{conformism}. 
The effects of conformity and non-conformity on cultural evolution have been studied with both models \citep{anticonformity,Denton2021,conformity} and experiments \citep{negativeFrequency}.

%\paragraph{Prestige.}
Prestige is defined as the ``widespread respect and admiration felt for someone or something on the basis of a perception of their achievements or quality'' (New Oxford American Dictionary).
Therefore, prestige does not directly signify success, although it may imply it--making it an indirect bias.
\citet[Ch.~8]{evolutionBook} and \citet{complexityPaper} suggest that prestige biases are probably more common in humans than success biases.
\citet[Ch. 8]{evolutionBook} add that maladaptive traits may spread widely in a population if indirect biases are strong enough.
They suggest that such biases could lead to a runaway process caused by a cultural equivalent of sexual selection \citep{sexualSelectionBook}.
On the other hand, \citet{fijian_social_bias} suggest that prestige biases, over generations, can lead to cultural adaptations, and that although prestige can lead to maladaptive traits spreading in the population, it can also accelerate the spread of adaptive traits.
Prestige is often mentioned in the cultural evolution literature, but seldom modeled.

%\paragraph{Influence bias.}
Today, social media provides an easy way to estimate the social and cultural influence individuals have over others, and therefore may have an effect on decision making. Online social networks such as \emph{Facebook} and \emph{Instagram} are known to affect the influence of individuals \citep{social_influence,social_media,influence_analysis}, and specific marketing practices were invented to capitalize on this effect \citep{facebook_marketing}.
Here, we model such influence as an indirect bias in cultural transmission, in which the choice of a role-model depends on the choices made by other individuals that have already chosen a role-model. 
Such a bias has been demonstrated by \citet{prestige_cultural_learning} in 3-4 year old children.

Influence bias is a context-based bias, rather than a content-based bias--it does not depend on the phenotype but rather on the number of copiers that already copied each role-model, which may be easier and more accurate to estimate compared to success.
Influence bias differs from frequency biases such as conformity, which depend on the frequency of a trait in the population or in a sample of role-models rather than the social dynamics of copying.

Here, we define a stochastic model of cultural transmission with prestige bias that combines both success and influence biases, find analytic approximations for this model, and analyze its dynamics. We find approximations for the probability and time for fixation of a `successful' phenotype (i.e., that is favored by success bias). Comparing these approximations to Kimura's approximations for the fixation of a favorable allele \citep{kimura,kimura_average}, we demonstrate that success and influence bias play the role of natural selection and genetic drift, respectively.

@@

\paragraph{Conclusions.}
Here, we presented support for the existence of prestige biases in cultural transmission, including the commonly studied success bias. We introduced an influence bias, which has so far received less attention. We believe the combination of these biases may be a strong driving force in cultural evolution. 
We show that the relative effect of the two biases changes the effective population size such that the increasing the weight individuals give to influence bias, decreases the effective population size. 
We also find a clear advantage to individuals that can choose the relative weight of the two biases.

>>>>

Cultural transmission may combine vertical transmission, in which parents transmit to their offspring; oblique transmission, in which adults (teachers, leaders, and even strangers) transmit to unrelated offspring; and horizontal transmission, in which peers from the same age cohort transmit to one another \citep{transmissionVectorsBook,transmissionVectors}.
It has been demonstrated that non-vertical cultural transmission can maintain maladaptive traits, which can be beneficial in changing environments \citep{changeEnv}.

Transmission biases may cause a cultural trait to have a disproportionate probability to be transmitted compared to its frequency in the population.
A common bias in the literature is \textit{success bias}, in which individuals prefer to copy from role models that demonstrate success in some activity, such as fishing, growing yams, and using medicinal plants \citep{fijian_social_bias}, and hunting \citep{Mesoudi2008}, and it can increase the probability of learning a successful trait \citep{Borofsky2022}. 
Indeed, in a tournament between learning strategies conducted by \citet{strategiesPaper}, most winner strategies included a mixture of success-biased social learning and individual learning, implying that success-biased learning is a good strategy, but that by itself it is not enough to best other strategies, even when success is measured accurately.

\citet[Ch.~5]{evolutionBook} have suggested that the evaluation of success can be divided into three groups: \emph{direct bias}, \emph{indirect bias} and \emph{frequency-dependent bias}.
A direct bias occurs when one phenotype is more attractive than other phenotypes, and is evaluated by \emph{directly} testing the trait.
For example, an individual observing a ping-pong match can attempt the observed paddle grips to determine which grip is better.
A frequency-dependent bias occurs when the probability to copy a phenotype is higher or lower than the frequency of the phenotype among demonstrators. 
For example, suppose the common paddle grip is used by 60\% of the demonstrators; if the this grip is adopted by 80\% of copiers, then transmission is under positive frequency bias, also called \emph{conformity}; if it is adopted by 40\% of copiers, then transmission is under negative frequency bias, or \emph{non-conformity} \citep{conformism}. 
The effects of conformity and non-conformity on cultural evolution have been studied with both models \citep{anticonformity,Denton2021,conformity} and experiments \citep{negativeFrequency}.
An indirect bias occurs when an individual uses one phenotype to determine the attractiveness of another phenotype.
For example, an observer may copy the paddle grip of the ping-pong player who scored more points in the match, thus indirectly evaluating the grip by the points scored.
This, however, may cause mismatches between the copied trait and the rest of the cultural or genetic repertoire of the individual \citep{Kolodny2022}.
Furthermore, \citet[Ch. 8]{evolutionBook} have suggested that maladaptive traits may spread widely in a population if indirect biases are strong enough by a runaway process caused by a cultural equivalent of sexual selection \citep{sexualSelectionBook}.
For example, \citet{cooperation} have shown that helping behaviors can evolve due to horizontal transmission bias even without any benefit to the recipient, or when the benefit is much larger than the cost.

\citet{fijian_social_bias} have studied such indirect biases, which they call cross-domain success bias, or in short, \emph{prestige} (e.g. great fishermen may be chosen as role-models for growing yams). They have suggested that prestige biases, over generations, can lead to cultural adaptations, and that although prestige can lead to maladaptive traits spreading in the population, it can also accelerate the spread of adaptive traits.
\citet[Ch.~8]{evolutionBook} and \citet{complexityPaper} have suggested that prestige biases are probably more common in humans than success biases, and prestige is often mentioned in the cultural evolution literature, however there are few models of prestige bias.

A broader definition of prestige is the ``widespread respect and admiration felt for someone or something on the basis of a perception of their achievements or quality'' (New Oxford American Dictionary).
Indeed, \citet{prestige_cultural_learning} have defined prestige bias as ``a tendency to learn from individuals to whom others have preferentially attended, learned or deferred'', and demonstrated its occurrence in in 3-4 year old children.
To distinguish this form of indirect bias from other definitions of prestige, we call it \emph{influence bias}. This is an important distinction, as influence is a context-based bias, rather than a content-based bias: it does not depend on the phenotype itself but rather on the number of copiers that already copied each role-model, which may be easier and more accurate to estimate compared to success. Influence bias differs from frequency biases such as conformity, which depend on the frequency of a trait in the population or in a sample of role-models, rather than the social dynamics of copying.
We model influence as an indirect bias in cultural transmission, in which the choice of a role-model depends on the choices made by other individuals that have already chosen a role-model.

In contemporary human society, social media makes it especially easy to estimate the social and cultural influence individuals have over others, and therefore can have an effect on decision making. Online social networks such as \emph{Facebook} and \emph{Instagram} are known to affect the influence of individuals \citep{social_influence,social_media,influence_analysis}, and specific marketing practices were invented to capitalize on this effect \citep{facebook_marketing}.

In the following, we develop a stochastic model of cultural transmission with that combines both success and influence biases, find analytic approximations for this model, and analyze its dynamics. We find approximations for the probability and time for fixation of a `successful' phenotype (i.e., that is favored by success bias). Comparing these approximations to Kimura's approximations for the fixation of a favorable allele \citep{kimura,kimura_average}, we demonstrate that success and influence bias play the role of natural selection and genetic drift, respectively.

@@

\paragraph{Conclusions.}
Here, we studied a model of cultural evolution under two transmission biases: the commonly studied success bias, together with influence bias, which has so far received less attention. We found approximations for this complex dynamics. We then showed that success bias affects the evolutionary dynamics much like natural selection does, whereas influence bias has a similar effect to random genetic drift. We also find a clear advantage to individuals that can choose the relative weight of the two biases.
